SU($n$) ferromagnets near the Heisenberg critical point

### 2020-04-09 collective_sun/sun_perturbation.tex
add sketch to explain permutation operator

Last but not least, SU($n$) symmetries appear naturally in the physics
of atomic, molecular, and optical (AMO)
systems\cite{gorshkov2010twoorbital, beverland2016realizing,
  cazalilla2014ultracold, taie2012su, hofrichter2016direct,
  cappellini2014direct, scazza2014observation, zhang2014spectroscopic,
  goban2018emergence, perlin2019effective} through the independence of
atomic orbital and interaction parameters on the $n$ nuclear spin
states of individual atoms, with e.g.~$n=10$ in the case of
${}^{87}$Sr.  As a result, experiments can directly probe the SU($n$)
Hubbard model, leading to experimental observations of SU($n$) Hubbard
phases and phase transitions\cite{taie2012su, hofrichter2016direct},
two-orbital SU($n$) magnetism\cite{cappellini2014direct,
  scazza2014observation, zhang2014spectroscopic}, and multi-body
SU($n$)-symmetric interactions\cite{goban2018emergence,
  perlin2019effective}.  The tremendous theoretical significance of
SU($n$) symmetries makes it all the more exciting that they appear in
experimental platforms with exquisite degrees of microscopic control,
thereby enabling the detailed study of these symmetries' consequences
for fundamental questions in physics, as well as their practical use
in technological applications.  SU(2)-symmetric interactions, for
example, can play an important role in the development of quantum
sensors that make use of entanglement to surpass classical limits on
measurement precision\cite{he2019engineering}\red{[cite XXZ squeezing
  paper]}.  The exciting prospect of similarly exploiting larger
SU($n$) symmetries to achieve a technological advantage is still an
unexplored avenue of research with enormous potential.

@@

In the following section, \ref{sec:bare_sun}, we introduce the SU($n$)
Heisenberg model, finding a convenient way to express its Hamiltonian
in terms of permutation operators that are agnostic to the value of
$n$.  In Section \ref{sec:breaking_sun}, we consider the introduction
of general spin operators, such as inhomogeneous external fields, that
weakly break SU($n$) symmetry.  We sketch out a derivation of the
effective Hamiltonian induced by these terms through second order in
perturbation theory, deferring more detailed calculations to
Appendices \ref{sec:eigenstates} and \ref{sec:operator_product}.
Finally, we discuss a numerical method to simulate dynamics within
manifolds of weak permutational symmetry breaking in Section
\ref{sec:shell_model}, before concluding in Section
\ref{sec:conclusions}.  Wherever possible, we will fall back to the
broadly familiar case of SU(2) in order to facilitate an understanding
of the general result for SU($n$).

@@

Before getting to the full SU($n$) Heisenberg model, we first revisit
the more common case of SU(2).  The SU(2) Heisenberg model consists of
$N$ two-level spins on a lattice, with spin-aligning interactions
\begin{align}
  H_0 = \sum_{\substack{\t{spin pairs}\\\set{i,j}}}
  h_{ij} \m s_i \c\m s_j,
  \label{eq:H_SU2}
\end{align}
where
$\m s_i\equiv\p{\sigma_\x^{(i)},\sigma_\y^{(i)},\sigma_\z^{(i)}}/2$ is
a spin vector defined with Pauli operators $\sigma_\alpha^{(i)}$
acting on the spin at lattice site $i$.  In the traditional language
of spin physics, each term of this Hamiltonian essentially imposes an
energy penalty of $h_{ij}$ to the formation of a spin 0 singlet state
$\ket{\up\dn}-\ket{\dn\up}$ between spins $i$ and $j$, favoring spin
triplet states with total spin 1.  Another interpretation of the
Hamiltonian in \eqref{eq:H_SU2} is illuminated by the observation that
the two-spin operator $\m s_i\c\m s_j$ is, up to an overall constant,
essentially the two-qubit SWAP operation $\Pi_{ij}$.  We will also
refer to $\Pi_{ij}$ as {\it permutation operator}, because its action
on the Hilbert space $\H\equiv\bigotimes_{k=1}^N\H_k$ of $N$ spins
with individual Hilbert spaces $\H_k$ is to permute the tensor factors
of $\H$ associated with spins $i$ and $j$.  Up to identity terms with
no physical consequence, we then have
\begin{align}
  H_0 = \sum_{\substack{\t{spin pairs}\\\set{i,j}}} h_{ij} \Pi_{ij},
  &&
  \Pi_{ij} \equiv \sum_{\substack{\t{spin states}\\\mu,\nu}}
  s_{\mu\nu}^{(i)} s_{\nu\mu}^{(j)}
  = \sum_{\substack{\t{spin states}\\\mu,\nu}} \op{\mu\nu}{\nu\mu}_{ij},
  \label{eq:H_0}
\end{align}
% todo: insert visualization of SWAP operation
where $s_{\mu\nu}^{(i)}\equiv\op{\mu}{\nu}_i$ is a spin-transition
operator for spin $i$, and $\mu,\nu\in\set{\up,\dn}$ index the states
of a single spin.  In addition to providing a nice interpretation of
the Hamiltonian $H_0$ as essentially swapping the states of spins $i$
and $j$ at a rate $h_{ij}$, the definition of the SU(2)-symmetric
Heisenberg model in in \eqref{eq:H_0} is straightforward to generalize
to SU($n$) with arbitrary $n$.  That is, rather than having
$\mu,\nu\in\set{\up,\dn}$, we simply let
$\mu,\nu\in\set{1,2,\cdots,n}$ index orthogonal states (sometimes
called {\it colors} or {\it flavors}) of an $n$-level spin.

@@

>>>>

Last but not least, SU($n$) symmetries appear naturally in the physics
of atomic, molecular, and optical (AMO)
systems\cite{gorshkov2010twoorbital, beverland2016realizing,
  cazalilla2014ultracold, taie2012su, hofrichter2016direct,
  cappellini2014direct, scazza2014observation, zhang2014spectroscopic,
  goban2018emergence, perlin2019effective} through the independence of
atomic orbital and interaction parameters on the $n$ nuclear spin
states of individual atoms, with e.g.~$n=10$ in the case of
${}^{87}$Sr.  As a result, experiments can directly probe the SU($n$)
Hubbard model, leading to experimental observations of SU($n$) Hubbard
phases and phase transitions\cite{taie2012su, hofrichter2016direct},
two-orbital SU($n$) magnetism\cite{cappellini2014direct,
  scazza2014observation, zhang2014spectroscopic}, and multi-body
SU($n$)-symmetric interactions\cite{goban2018emergence,
  perlin2019effective}.  The tremendous theoretical significance of
SU($n$) symmetries makes it all the more exciting that they appear in
experimental platforms with exquisite degrees of microscopic control,
thereby enabling the detailed study of these symmetries' consequences
for fundamental questions in physics, as well as their practical use
in technological applications.  SU(2)-symmetric spin interactions, for
example, can play an important role in the development of quantum
sensors that make use of entanglement to surpass classical limits on
measurement precision\cite{he2019engineering}\red{[cite XXZ squeezing
  paper]}.  The exciting prospect of similarly exploiting larger
SU($n$) symmetries to achieve a technological advantage is still an
unexplored avenue of research with enormous potential.

@@

In the following section, \ref{sec:bare_sun}, we introduce the SU($n$)
Heisenberg model, finding a convenient way to express its Hamiltonian
in terms of SWAP operations that are agnostic to the value of $n$.  In
Section \ref{sec:breaking_sun}, we consider modifying the Hamiltonian
of the bare Heisenberg model by the addition of spin operators,
corresponding to e.g.~external fields or additional interactions, that
weakly break SU($n$) symmetry.  We sketch out a derivation of the
effective Hamiltonian induced by these terms through second order in
perturbation theory, deferring more detailed calculations to
Appendices \ref{sec:eigenstates} and \ref{sec:operator_product}.
Finally, we discuss a numerical method to simulate dynamics within
manifolds of weak permutational symmetry breaking in Section
\ref{sec:shell_model}, before concluding in Section
\ref{sec:conclusions}.  Wherever possible, we will fall back to the
broadly familiar case of SU(2) in order to facilitate an understanding
of the generalizations to SU($n$).

@@

Before getting to the full SU($n$) Heisenberg model, we first revisit
the more common case of SU(2).  The SU(2) Heisenberg model consists of
$N$ two-level spins on a lattice, with spin-aligning interactions
\begin{align}
  H_0 = \sum_{\substack{\t{spin pairs}\\\set{i,j}}}
  h_{ij} \m s_i \c\m s_j,
  \label{eq:H_SU2}
\end{align}
where
$\m s_i\equiv\f12\p{\sigma_\x^{(i)},\sigma_\y^{(i)},\sigma_\z^{(i)}}$
is a spin vector defined with Pauli operators $\sigma_\alpha^{(i)}$
acting on the spin at lattice site $i$.  In the standard language of
spin physics, each term of this Hamiltonian essentially imposes an
energy penalty of $-h_{ij}$ to the formation of a spin 0 singlet state
$\ket{\up\dn}-\ket{\dn\up}$ between spins $i$ and $j$, favoring the
formation of spin triplet states with total spin 1.  Another
interpretation of the Hamiltonian in \eqref{eq:H_SU2} is illuminated
by the observation that the two-spin operator $\m s_i\c\m s_j$ is, up
to an overall constant, essentially the two-qubit SWAP operation
$\Pi_{ij}$.  We will also refer to $\Pi_{ij}$ as {\it permutation
  operator}, because its action on the Hilbert space
$\H\equiv\bigotimes_{k=1}^N\H_k$ of $N$ spins with individual Hilbert
spaces $\H_k$ is to permute the tensor factors of $\H$ associated with
spins $i$ and $j$ (see Figure \ref{fig:sketch_spins}).  Up to identity
terms with no physical consequence, we then have
\begin{align}
  H_0 = \sum_{\substack{\t{spin pairs}\\\set{i,j}}} h_{ij} \Pi_{ij},
  &&
  \Pi_{ij} \equiv \sum_{\substack{\t{spin states}\\\mu,\nu}}
  s_{\mu\nu}^{(i)} s_{\nu\mu}^{(j)}
  = \sum_{\substack{\t{spin states}\\\mu,\nu}} \op{\mu\nu}{\nu\mu}_{ij},
  \label{eq:H_0}
\end{align}
where $s_{\mu\nu}^{(i)}\equiv\op{\mu}{\nu}_i$ is a spin-transition
operator for spin $i$, and $\mu,\nu\in\set{\up,\dn}$ index the states
of a single spin.  In addition to providing a nice interpretation of
the Hamiltonian $H_0$ as essentially swapping the states of spins $i$
and $j$ at a rate $h_{ij}$, the definition of the SU(2)-symmetric
Heisenberg model in in \eqref{eq:H_0} is straightforward to generalize
to SU($n$) with arbitrary $n$.  That is, rather than having
$\mu,\nu\in\set{\up,\dn}$ index two spin states, we simply let
$\mu,\nu\in\set{1,2,\cdots,n}$ index the orthogonal states (sometimes
called {\it colors} or {\it flavors}) of an $n$-level spin.

\begin{figure}
  \centering
  \includegraphics[valign=m]{sketch_spins.pdf}
  $\stackrel{\displaystyle\Pi_{ij}}
  {\displaystyle\longleftrightarrow}$
  \includegraphics[valign=m]{sketch_spins_swap.pdf}
  \caption{The permutation operator $\Pi_{ij}$ swaps the states of
    spins $i$ and $j$, which in the case of SU(2) can be represented
    by magnetization vectors in 3-dimensional space.  In the case of
    SU($n$), these vectors live in a higher-dimensional space without
    a clear geometric interpretation, but the interpretation of the
    permutation operator $\Pi_{ij}$ remains the same.}
  \label{fig:sketch_spins}
\end{figure}

@@

\vspace{3cm}
